An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-04-22 EconGrowth.Rnw
Converted the triangle plots to have a columnar structure rather than a row structure. This allows the x-axis parameter to be the same for any column of panels. This, in turn, allows x axis labels to be moved out of the strip and onto the x axis!

\item{Heun: Fix warnings of the form ``Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562'' in the code that generates the Cobb-Douglas with energy fits.}

@@

The key energy parameters for the Cobb-Douglas and CES models are $\gamma$, $\delta$, and $\sigma$. When the 95\% confidence interval for $\gamma$ and $\delta$ extends to 0.0 and 1.0, respectively, we cannot say whether including energy in the production function is helpful. If the value of $\gamma$ is close to 0.0, and the 95\% confidence interval does not extend to 0.0, we know that energy does not play an important role in producing economic output for the Cobb-Douglas model. If the value of $\delta$ is close to 1.0 and the 95\% confidence intervale does not extend to 1.0, we know that energy does not play an important role in producing economic output for the CES model.

The following figure shows these relationships.

<<CIvsParam_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="95\\% Confidence interval width vs. parameter value for key energy parameters ($\\gamma$, $\\delta$, and $\\sigma$) of the Cobb-Douglas and CES models.">>=
CIvsParamLattice(countryAbbrevScaling=0.6)
@

>>>>

\item{Heun: Fix warnings of the form \texttt{Warning:  step factor 0.000488281 reduced below ’minFactor’ of 0.000976562} in the code that generates the Cobb-Douglas with energy fits.}

@@

The key energy parameters for the Cobb-Douglas and CES models are $\gamma$, $\delta$, and $\sigma$. The parameter $\gamma$ in the Cobb-Douglas model (Equation \ref{eq:CD_With_Energy}) represents the importance of energy for economic growth. When $\gamma$ is small (close to 0), energy is not a significant contributor to economic growth for an economy. Conversely, when $\gamma$ is large (close to 1), energy is the primary contributor to economic growth. However, if the 95\% confidence interval for $\gamma$ for any given economy extends to 0.0, we cannot say with high confidence that energy makes a meaningful contribution to economic growth for that economy. On a graph of the width of the 95\% confidence interval for 95\% confidence interval on $\gamma$ ($CI_{95\%,\gamma}$) vs. $\gamma$, where each economy is represented by a point, the region where the 95\% confidence interval on $\gamma$ crosses 0 is an upper-left triangle, above and to the left of a line that connects points (0, 0) and (1, 2) on the graph. If $\gamma$ for a given economy is below and to the right of the line connecting (0, 0) and (1, 2), we have high confidence that we can say whether or not energy is a significant contributor to economic growth. Thus, only when $\gamma$ is below the line and small (i.e., close to 0) for an economy, can we say with confidence that energy is an insignificant contributor to economic growth. Only when $\gamma$ is below the line and large (i.e., close to 1) for an economy can we say that energy is a significant contributor to economic growth, according to the Cobb-Douglas model.

The left column of graphs in Figure \ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\gamma}$ vs. $\gamma$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that most countries are above and to the left of the line, while only a few are below and to the right.

For the CES model (Equation \ref{eq:micEconCES_model}), the parameter $\delta$ indicates that importance of energy for economic growth. When $\delta$ is close to 1, energy is an insignificant contributor to economic growth, and when $\delta$ is close to 0, energy is the primary contributor to economic growth for an economy. However, if the 95\% confidence interval for $\delta$ extends to 1, we cannot say with high confidence that energy makes a meaningful contribution to economic growth for that economy. On a graph of $\delta$ confidence interval width ($CI_{95\%,\delta}$) vs. $\delta$, where each economy is represented by a point, the region where the 95\% confidence interval for $\delta$ crosses 1 is an upper-right triangle, above and to the right of a line that connects points (1, 0) and (0, 2) on the graph. If $\delta$ for a given economy is below and to the left of the line connecting (1, 0) and (0, 2), we have high confidence that we can say whether or not energy is a significant contributor to economic growth. Thus, only when $\delta$ is below the line and large (i.e., close to 1) for an economy, can we say with confidence that energy is an insignificant contributor to economic growth. Only when $\delta$ is below the line and small (i.e., close to 0) for an economy can we say that energy is a significant contributor to economic growth, according to the CES model.

The middle column of graphs in Figure \ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\delta}$ vs. $\delta$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that most countries are above and to the right of the line, while only a few are below and to the left.

For the CES model (Equation \ref{eq:micEconCES_model}), the parameter $\sigma$ indicates whether capital stock and labor ($kl$) and energy ($e$) are substitutes ($\sigma >$ 1) or complements ($\sigma <$ 1). For countries whose value of $\sigma$ is below and to the right of a line connecting (0, 0) and (2, 1) on a graph of $CI_{95\%,\sigma}$ vs. $\sigma$, we can tell (with 95\% confidence) whether ($kl$) and ($e$) are substitutes or complements. For economies whose value of $\sigma$ is above and to the left of a line connecting (0, 0) and (2, 1), we cannot say with 95\% confidence whether ($kl$) and ($e$) are substitues or complements. Conversely, for economies whose value of $\sigma$ is below and to the right of a line connecting (0, 0) and (2, 1), we are able to say with 95\% confidence whether ($kl$) and ($e$) are substitues or complements.

The right column of graphs in Figure \ref{fig:CIvsParam_Lattice_Graph} shows $CI_{95\%,\sigma}$ vs. $\sigma$ for the three quantifications of energy (thermal energy, $q$, exergy, $x$, and useful work, $u$). We note that where we can tell one way or another, energy ($e$) appears to complement capital stock and labor ($kl$).

<<CIvsParam_Lattice_Graph, fig.width=ninePanelGraphWidth, fig.height=ninePanelGraphHeight, fig.pos="H", fig.cap="95\\% Confidence interval width vs. parameter value for key energy parameters ($\\gamma$, $\\delta$, and $\\sigma$) of the Cobb-Douglas and CES models.", eval=TRUE>>=
CIvsParamLattice(countryAbbrevScaling=0.6)
@

*********** Martin: Need interpretation here. *******************
